---
title: "SyntheticData"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SyntheticData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(densityratio)
library(synthpop)
library(dplyr)
```

```{r}
# data <- read.csv(url("https://raw.githubusercontent.com/avehtari/ROS-Examples/master/KidIQ/data/kidiq.csv"))

data <- readRDS("kids.rds") |> select(-mom_hs)
  mutate(mom_hs = as.factor(mom_hs))

mysyn <- syn(data)
```

```{r}
densratio <- ulsif(mysyn$syn, data)
class(mysyn$syn)
```


# Introduction

Density ratios have many usages.
One is covariate-shift.
One is synthetic data.
In this vignette, we demonstrate how to use it to create synthetic data.

# Data

Brief data description and source. Included in the vignette.

# Synthetic data

We use the package synthpop. We start with a deliberately wrong model.
We examine plot_univariate(), and plot_bivariate(), and plot. We check a summary.

We use the defualt model, which gives an appropriate model.

# Conclusion

'Stupid' example.


